SmartCheck

### 2012-12-05 paper/paper.tex
paper mods.

  newVals (SubVal a) = 
    sequence (replicate maxTries s)
    where s =     sizedArbitrary a 
              >>= return . SubVal

test :: SubTypes a 
     => a -> Int -> [SubVal] 
     -> (a -> Property) -> Maybe a
test cex idx vs prop = go vs
  where
  go []      = Nothing
  go (v:vs') = 
    let cex' = replace cex idx v in
    if pass prop cex' then go vs' 
      else Just cex'
  \end{lstlisting}
  \caption{Counterexample reduction algorithm\label{fig:reduction}}
\end{figure}

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes a counterexample \ttp{cex} and the property \ttp{prop}.  At
the beginning of execution, we assume that a counterexample has been obtained
somehow; our implementation relies on QuickCheck as a ``front-end'' to generate
a counterexample that the algorithm automatically shrinks.  The reduction begins
at the first sub-value of \ttp{cex}, \ttp{v}.  When the index \ttp{idx} becomes
out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.  Otherwise, a
list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a  
\end{code}
%
generates a new value \ttp{v'} having the same type as \ttp{v} and strictly
smaller (with respect to the \ttp{size} method) than \ttp{v}.
\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, we assume one is generated locally).  The constant \ttp{maxTries} is
the maximum number of tries to discover a new counterexample by replacing
\ttp{v} in \ttp{cex} and testing it.  The result of \ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
holds if \ttp{cex'} satisfies the property \ttp{prop}.  If no failure is found,
we move to the next sub-value of \ttp{cex} and continue.  However, if a new
smaller counterexample \ttp{cex'} is found, we start a new breadth-first
traversal of \ttp{cex'}, attempting to shrink it further.

@@

\paragraph{Sum-type hypothesis}
First, recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the 0th sub-value.  The 0th sub-value of \ttp{cex} is
\ttp{cex} itself.  No invariant of the algorithm would be violated by beginning
with the 0th sub-value; recall that it would generate strictly smaller values
than \ttp{cex}.  However, we begin with a strict sub-value because of a
hypothesis we refer to as the \emph{sum-type hypothesis}.  The hypothesis is
that in the search space of possible values, if a counterexample of the form
\ttp{C}~$\vec{x}$, where \ttp{C} is some constructor and $\vec{x}$ are its
arguments, is discovered, then it is more likely to find another ``nearby''
counterexample of the form \ttp{C}~$\vec{y}$, where $\vec{y}$ are different
arguments to \ttp{C}.  For example, consider a property about unintended
variable capture over a language's parse tree represented by a sum type with
constructors for module imports, function definitions, and global-variable
assignments, respectively.  A function definition counterexample can only be
reduced to smaller function definition counterexamples, the only construct in
which variable capture is possible.

@@

\paragraph{Counterexample search}
We make two optimizations in searching for counterexamples.  First, just like
QuickCheck's \ttp{arbitrary} method does, our \ttp{sizedArbitrary} function
successfully larger counterexamples when generating new values with which to
replace a sub-value.

Furthermore, we can instrument a low-cost shortcut check in the algorithm.
Sometimes, a counterexample fails a property due to a sub-value nested deep
inside the counterexample.  For example, consider a simple calculator language
containing constant \ttp{Int}s, addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp  
\end{code}
%
\noindent
and consider an evaluator for the language that returns \ttp{Nothing} if
division-by-zero is encountered and \ttp{Just} the value otherwise:
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) = 
  let e = eval e1 in 
  if e == Just 0 then Nothing 
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Suppose we restrict the set of programs with the following predicate:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =    divSubTerms e0 
                            && divSubTerms e1
divSubTerms (Div e0 e1)   =    divSubTerms e0 
                            && divSubTerms e1  
\end{code}
%
and then test the property
%
\begin{code}
div_prop e = divSubTerms e ==> eval e /= Nothing  
\end{code}
%
\noindent
The property \ttp{div\_prop} turns out to be false; for example, 
%
\begin{code}
Div (C 1) (Div (C 0) (C 1))
\end{code}
%
\noindent
is a counterexample.  In the predicate case
%
\begin{code}
divSubTerms (Div _ (C 0)) = False  
\end{code}
%
\noindent
we forgot to recur on the divisor.  However, a typical counterexample returned
by QuickCheck, without \ttp{shrink} implemented, is as follows:
%
\begin{quote}
\ttp{Div (Div (C (-22)) (Div (C 2) (C 13))) (Div (Div (Div (C (-11)) (C 14))
  (Add (Div (C (-4)) (C (-3))) (Div (C (-25)) (C (-25))))) (Add (Add (C 22) (Div
  (C (-30)) (C (-14)))) (C 13)))}
\end{quote}
%
\noindent
To butcher an expression, finding the cause of failure is like searching for a
divide-by-zero in the haystack!  But it turns out that the culprit is a small
sub-value:
%
\begin{code}
Div (C (-22)) (Div (C 2) (C 13))  
\end{code}
%

Generalizing this scenario, during the reduction algorithm's breadth-first
search through a counterexample \ttp{cex}'s sub-values, we may happen upon a
sub-value \ttp{cex'} that has the same type as \ttp{cex} and fails the property
(while passing any preconditions).  In this case, we can return \ttp{cex'}
directly, and rerun the reduction algorithm on \ttp{cex'}.

\todo{violates sum-type hypothesis but cheap---not testing generated values,
  just direct subvalues.  Call this the sub-value hypothesis: direct sub-values
  of a counterexample may also be a counterexample.}

>>>>

newVals :: SubVal -> IO [SubVal]
newVals (SubVal a) = 
  sequence (replicate maxTries s)
  where s =     sizedArbitrary a 
            >>= return . SubVal

test :: SubTypes a 
     => a -> Int -> [SubVal] 
     -> (a -> Property) -> Maybe a
test cex idx vs prop = go vs
  where
  go []      = Nothing
  go (v:vs') = 
    let cex' = replace cex idx v in
    if pass prop cex' then go vs' 
      else Just cex'
  \end{lstlisting}
  \caption{Counterexample reduction algorithm.\label{fig:reduction}}
\end{figure}

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes a counterexample \ttp{cex} and the property \ttp{prop}.  At
the beginning of execution, we assume that a counterexample has been obtained
somehow; our implementation relies on QuickCheck as a ``front-end'' to generate
a counterexample that the algorithm automatically shrinks.  The reduction begins
at the first sub-value of \ttp{cex}, \ttp{v}.  When the index \ttp{idx} becomes
out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.  Otherwise, a
list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a  
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and strictly
smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just like
QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates successfully
larger counterexamples when generating new values with which to replace a
sub-value.

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, we assume one is generated locally).  The constant \ttp{maxTries} is
the maximum number of tries to discover a new counterexample by replacing
\ttp{v} in \ttp{cex} and testing it.  The result of \ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  If no failure is found,
we move to the next sub-value of \ttp{cex} and continue.  However, if a new
smaller counterexample \ttp{cex'} is found, we start a new breadth-first
traversal of \ttp{cex'}, attempting to shrink it further.

@@

\paragraph{Sum-type counterexample hypothesis}
First, recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the 0th sub-value.  The 0th sub-value of \ttp{cex} is
\ttp{cex} itself.  No invariant of the algorithm would be violated by beginning
with the 0th sub-value; recall that it would generate strictly smaller values
than \ttp{cex}.  However, we begin with a strict sub-value because of a
hypothesis we refer to as the \emph{sum-type counterexample hypothesis}.  The
hypothesis is that in the search space of possible values, if a counterexample
of the form \ttp{C}~$\vec{x}$, where \ttp{C} is some constructor and $\vec{x}$
are its arguments, is discovered, then it is more likely to find another
``nearby'' counterexample of the form \ttp{C}~$\vec{y}$, where $\vec{y}$ are
different arguments to \ttp{C}.  For example, consider a property about
unintended variable capture over a language's parse tree represented by a sum
type with constructors for module imports, function definitions, and
global-variable assignments, respectively.  A function definition counterexample
can only be reduced to smaller function definition counterexamples, the only
construct in which variable capture is possible.

@@

\paragraph{Sub-value counterexample hypothesis}
Sometimes, a counterexample fails a property due to a sub-value nested deep
inside the counterexample.  The rest of the value is irrelevant.  We call this
the \emph{sub-value counterexample hypothesis.}  Thus, one way to efficiently
search the space of potential counterexamples is to test a counterexample's
(well-typed) sub-values.

For example, consider a simple calculator language containing constant
\ttp{Int}s, addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp  
\end{code}
%
\noindent
and consider an interpreter for the language that returns \ttp{Nothing} if
division-by-zero is encountered and \ttp{Just} the value otherwise:
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) = 
  let e = eval e1 in 
  if e == Just 0 then Nothing 
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Suppose we restrict the set of programs with the following predicate:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =    divSubTerms e0 
                            && divSubTerms e1
divSubTerms (Div e0 e1)   =    divSubTerms e0 
                            && divSubTerms e1  
\end{code}
%
and then test the property
%
\begin{code}
div_prop e = divSubTerms e ==> eval e /= Nothing  
\end{code}
%
\noindent
The property \ttp{div\_prop} turns out to be false; for example, 
%
\begin{code}
Div (C 1) (Div (C 0) (C 1))
\end{code}
%
\noindent
is a counterexample.  In the predicate case
%
\begin{code}
divSubTerms (Div _ (C 0)) = False  
\end{code}
%
\noindent
we forgot to recur on the divisor.  However, a typical counterexample returned
by QuickCheck, without \ttp{shrink} implemented, is as follows:
%
\begin{quote}
\ttp{Div (Div (C (-22)) (Div (C 2) (C 13))) (Div (Div (Div (C (-11)) (C 14))
  (Add (Div (C (-4)) (C (-3))) (Div (C (-25)) (C (-25))))) (Add (Add (C 22) (Div
  (C (-30)) (C (-14)))) (C 13)))}
\end{quote}
%
\noindent
To butcher an expression, finding the cause of failure is like searching for a
divide-by-zero in the haystack!  But it turns out that the culprit is a small
sub-value:
%
\begin{code}
Div (C (-22)) (Div (C 2) (C 13))  
\end{code}
%

\begin{figure}
\begin{lstlisting}
reduce0 :: forall a . SubTypes a 
        => a -> (a -> Property) -> IO a
reduce0 cex prop = reduce' 1
  where
  reduce' idx = 
    case index cex idx of 
      Nothing -> return cex
      Just v  -> 
        case testHole v of
          Just v' -> reduce0 v' prop
          Nothing -> do
            vs <- newVals v
            case test cex idx vs prop of
              Nothing   -> 
                reduce' (idx+1)
              Just cex' -> 
                reduce0 cex' prop

  testHole (SubVal a) = 
    case cast a :: Maybe a of
      Nothing -> Nothing
      Just a' -> 
        if pass prop a' then Nothing
          else Just a'
\end{lstlisting}
  \caption{Reduction algorithm with the sub-value counterexample optimization.}
  \label{fig:reduce0}
\end{figure}

Generalizing this scenario, during the reduction algorithm's breadth-first
search through a counterexample \ttp{cex}'s sub-values, we may happen upon a
sub-value \ttp{cex'} that has the same type as \ttp{cex} and fails the property
(while passing any preconditions).  In this case, we can return \ttp{cex'}
directly, and rerun the reduction algorithm on \ttp{cex'}.  In
Figure~\ref{fig:reduce0}, we show an updated reduction algorithm, \ttp{reduce0}
that implements this optimization.  The function \ttp{testHole} tests the
current sub-value and if it fails the property, the function tests the sub-value
directly.
